Constructing a universe for the setoid model

### 2020-06-24 paper.tex
More paper writing

A central aspect of Type Theory is the coexistence of two notions of
equality. On the one hand definitional equality, the computational equality that
is built into the formalism. On the other hand propositional equality, the
internal notion of equality that is actually used to state and prove equational
theorems within the system. The precise balance between these two notions is at
the center of type theory research; however, it is generally understood that to
properly support formalization of mathematics in type theory, one should aim for
a definitional equality that is decidable and therefore machine-checkable, and a
notion of propositional equality that is as extensional as possible.

Two extensionality principles seem particularly desirable, in that they arguably
constitute the bare minimum for type theory to be comparable to set theory as a
foundational system for set-level mathematics, in terms of power and ergonomics.
One is function extensionality, according to which functions are equal if
point-wise equal. Another is propositional extensionality, equating all
propositions that are logically equivalent.

Extensional Type Theory (ETT) does support extensional reasoning to some degree,
thus seemingly being a convenient setting for formalization of
mathematics. However, equality reflection makes type-checking ETT terms
computationally unfeasible, if not impossible.

On the other hand, Intensional Type Theory has nice computational properties
that can make it more suitable for computer implementation, but as normally
defined [cite ML paper?] it lacks the amount of extensionality necessary to
effectively \emph{use} it.
%
It is known from model constructions that extensional principles like function
extensionality (funext) are consistent with ITT.
%
Moreover, ITT extended with uniqueness of identity proofs (UIP) and funext is
known to be as powerful as ETT [cite hofmann]. These principles can be added as
axioms, however destroying some computational properties of ITT such as
canonicity.

Ultimately, what is needed is a formulation of ITT that supports the
aforementioned extensionality principles, while retaining convenient
computational properties like canonicity.
%
Unfortunately, one can show that whenever identity types are defined according
to Martin-Loef's inductive definition, there is no hope to have canonicity and
funext coexist.
%
The first step towards a solution is to give up the idea of equality as a single
inductive definition. Instead, equality should be defined specifically for each
type former in the type theory. According to this principle, funext can be
justified by simply defining equality on function types to be point-wise
equality.

%% * One possible answer: setoids

Sets, or types, together with a propositional equivalence relation as their
notion of equality are known as \emph{setoids}. There are several definitions of
the setoid model of Type Theory, the first of which was constructed to show
consistency of funext for ITT [cite]. This consistency result is implied by the
observation that UIP and funext hold in the setoid model. Moreover, it was shown
by Altenkirch [cite] that is the model construction is carried out in a
metatheory with a universe of strict (definitionally proof-irrelevant)
propositions, it is possible to define a univalent universe of propositions,
therefore satisfying propositional extensionality. The setoid model thus
satisfies all the extensionality principles that we would like to extend ITT
with. The question is whether there exists a version of ITT that supports setoid
reasoning, and hence the forms of extensionality enables by it.

@@

This section describes $MLTT^{Prop}$, the metatheory in which we model
construction takes place. It is crucial to keep the metatheory as minimal as
possible.

$MLTT^{Prop}$ is an intensional MLTT with $\Pi, \Sigma, \textsf{Bool},
\mathbf{0}, \mathbf{1}$, and a universe of strict propositions $\mProp$. We also
have definitional $\eta$ rules for $\Pi, \Sigma$, and $\mathbf{1}$
types. Propositions can be lifted to types via $\textsf{Lift} : \mProp \to
\mType$. The universe of propositions is inhabited by false $\bot$ and
propositional truncation $\|\_\| : \mType \to \mProp$. We also require indexed
W-types, both in $\mType$ and $\mProp$:

@@

We also require the propositional versions of $\mathbf{1}$, $\Pi$, and $\Sigma$,
however these can be defined from their $\mType$ counterparts via truncation:

@@

\[
\infer{
  \begin{aligned}
    & | A | : | \GG | \to \mType \\
    & A^\sim : \{\gamma_0\ \gamma_1 : | \GG |\}
    \to | A | \gamma_0 \to | A | \gamma_1 \to \mProp \\
    & \textsf{coe} : \{\gamma_0\ \gamma_1 : | \GG |\} \to \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \\
    & \dots
  \end{aligned}
}{
  A : \Ty\ \GG
}
\]

@@

An inductive-recursive universe is given by a type of codes $\U$, and a family
$\El : \U \to \mType$ that assigns, to each code corresponding to some type, the
meta-theoretic type of its elements. The resulting definition is
inductive-recursive because the inductive type of codes is defined
simultaneously with the recursive function $\El$.

@@

\begin{align*}
  & \textsf{data}\ \U\ : \mType_1 \\
  & \quad \textsf{bool} : \U \\
  & \quad \textsf{pi} : (A : \U) \to (\El\ A \to \U) \to \U \\
  & \El\ \textsf{bool} = \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B) = (a : \El\ A) \to \El\ (B\ a)
\end{align*}

Note that $\U$ is defined to be in $\mType_1$ rather than $\mType$. This is
because... \todo{TODO add explanation}. This fact also implies that the
meta-theory must contain at least one universe.

Induction-recursion is arguably the nicest and more natural way to define
internal universes in Type Theory, however it is not always required.
%
We can translate basic instances of induction-recursion into inductive families
using the equivalence of $I$-indexed families of sets and sets over $I$,
i.e.\ $[I, \mType_i] \simeq \mType_i/I$ for $I : \mType_i$.

@@

Note that this construction gives rise to a universe in $\mType_1$, rather than
$\Set$, since the definition of $\U$ quantifies over all possible sets.

@@

To construct the universe of setoids in the setoid model, we first of all need
to define a type $\Uty : \Ty\ \GG$ for every $\GG : \Con$, and for every $A :
\Tm\ \GG\ \Uty$ a type $\mathds{E}l\ A : \Ty\ \GG$. Recalling Section TODO,
these are essentially record types made of several Since. components $\Uty$ is a
closed type, it requires the same data of a setoid; in particular, we need a set
of codes together with an equivalence relation reflecting equality of codes
%
\begin{align*}
  & \textsf{data} \ \setoidU : \mType_1 \\
  & \_\sim_\setoidU\_ : \setoidU \to \setoidU \to \mProp_1
  %% & \El : \setoidU \to \Set \\
  %% & \_\vdash\_\sim_\El\_ : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \Prop
\end{align*}
%
in addition to proofs that these are indeed equivalence relations
%
\begin{align*}
  & \textsf{refl}_\setoidU : (A : \setoidU) \to \equ{A}{A} \\
  & \textsf{sym}_\setoidU : \equ{A}{B} \to \equ{B}{A} \\
  & \textsf{trans}_\setoidU : \equ{A}{B} \to \equ{B}{C} \to \equ{A}{C}
\end{align*}

@@

Note that $\eqel{}{}{}$ is indexed over equality on the universe, because $\El$
is a setoid morphism, hence in particular it must respect the setoid equality of
$\setoidU$.
%
We also require data and proofs that make sure we get setoids out of $\El$:

@@

\todo{why is this inductive-inductive-recursive?}

@@

The reduction of the IIR universe to induction works, conceptually, as before.
%
However, in addition to $\ad{in-U}$ for defining $\U$, we also introduce a
family $\ad{in-U∼}$ of binary relations between types in the universe, from
which we then define $\_\sim_{\setoidU}\_$.
%
\begin{align*}
  & \textsf{data}\ \ad{in-U} : \Set \to \Set_1 \\
  & \textsf{data}\ \ad{in-U∼} : \{A\ A' : \Set\} \to \ad{in-U}\ A \to \ad{in-U}\ A' \to (A \to A' \to \Prop) \to \Set_1
\end{align*}

Note that $\ad{in-U}$ appears in the sort of $\ad{in-U∼}$. This is an instance
of \emph{induction-induction}, a schema that allows the definition of a type
mutually with other types that contain the first one in their signature [cit].

@@

The full definition of the universe is concluded with the remaining definitions,
like $\reflu, \reflel$, etc., which can be adapted from their IR counterparts
more or less straighforwardly. The final result does not use
induction-recursion, but it's nevertheless an instance of infinitary
induction-induction. The ability to define arbitrary, infinitary
inductive-inductive types clashes with our objective of keeping the metatheory
as small and simple as possible. The next step is therefore to reduce this
inductive-inductive universe to one that does not require (infinitary)
induction-induction.

>>>>

A central aspect of Type Theory is the coexistence of two notions of
equality. On the one hand definitional equality, the computational equality that
is built into the formalism. On the other hand ``propositional'' equality, the
internal notion of equality that is actually used to state and prove equational
theorems within the system. The precise balance between these two notions is at
the center of type theory research; however, it is generally understood that to
properly support formalization of mathematics in type theory, one should aim for
a notion of definitional equality that is decidable and therefore
machine-checkable, and a notion of propositional equality that is as as
extensional possible.

Two extensionality principles seem particularly desirable, in that they arguably
constitute the bare minimum for type theory to be comparable to set theory as a
foundational system for set-level mathematics, in terms of power and ergonomics.
One is function extensionality (funext), according to which functions are equal
if point-wise equal. Another is propositional extensionality (propext), equating
all propositions that are logically equivalent.

Extensional Type Theory (ETT) does support extensional reasoning to some degree,
and thus would seem a convenient setting for formalization of
mathematics. However, ETT's equality reflection makes type-checking ETT terms
computationally unfeasible, if not impossible.

On the other hand, Intensional Type Theory (ITT) has nice computational
properties that can make it more suitable for computer implementation, but as
normally defined [cite Martin-L\"of paper?] it lacks extensionality.
%
It is known from model constructions that extensional principles like funext are
consistent with ITT.
%
Moreover, ITT extended with uniqueness of identity proofs (UIP) and funext is
known to be as powerful as ETT [cite hofmann]. These principles can be added as
axioms, however destroying some computational properties of ITT such as
canonicity.

What we would like instead is a formulation of ITT that supports the
aforementioned extensionality principles, while retaining convenient
computational properties like canonicity and decidability of type checking.
%
Unfortunately, one can show that whenever identity types are defined according
to Martin-Loef's inductive definition, there is no hope to have canonicity and
funext coexist.
%
The first step towards a solution is to give up the idea of propositional
equality as single type family inductively generated by reflexivity. Instead,
equality should be \emph{specific} to each type former in the type theory, or in
other words, every type former should be introduced alongside an explanation of
what counts as equality for its elements. In this setting, function
extensionality can simply be introduced as the chosen notion of equality on
function types.

This conception of types together with their notion of equality goes back to the
notion of \emph{setoids} or \emph{Bishop sets}. A setoid is simply a set
equipped with a propositional equivalence relation. Setoids provide a quite
natural and useful semantic domain in which to interpret type theory. The first
setoid model was constructed to show consistency of funext for ITT [cite]. This
consistency result is implied by the observation that UIP and funext hold in the
setoid model. Moreover, it was shown by Altenkirch [cite] that is the model
construction is carried out in a metatheory with a universe of strict
(definitionally proof-irrelevant) propositions, it is possible to define a
univalent universe of propositions, therefore satisfying propositional
extensionality. The setoid model thus satisfies all the extensionality
principles that we would like to extend ITT with. The question is whether there
exists a version of ITT that supports setoid reasoning, and hence the forms of
extensionality enables by it.

@@

This section describes $MLTT^{Prop}$, the metatheory in which the model
construction described in the following sections takes place. One aspect of the
metatheory that we pay attention to is it's minimality, as we want to keep it as
simple as possible. We want to particularly avoid having to assume strong
definition schemas that go beyond inductive families (or, semantically, what can
be modeled with W types.)

$MLTT^{Prop}$ is an intensional type theory in the style of Martin-L\"of Type
Theory with $\Pi, \Sigma, \textsf{Bool}, \mathbf{0}, \mathbf{1}$, and a universe
of strict propositions $\mProp$. $\Sigma$ types are defined negatively by
pairing $\_,\_$ and projections $\pi_1, \pi_2$. We also have definitional $\eta$
rules for $\Pi, \Sigma$, and $\mathbf{1}$ types. Propositions can be lifted to
types via $\textsf{Lift} : \mProp \to \mType$. The universe of propositions is
inhabited by false $\bot$ and propositional truncation $\|\_\| : \mType \to
\mProp$. We also require indexed W-types, both in $\mType$ and $\mProp$:

@@

In addition to the proof-relevant types just listed, we will need the
propositional versions of $\mathbf{1}$, $\Pi$, and $\Sigma$; however these can
be defined from their $\mType$ counterparts via truncation. That is, given $P :
\mProp$ and $Q : P \to \mProp$:

@@

\[
\infer{
  \begin{aligned}
    & | A | : | \GG | \to \mType \\
    & A^\sim : \{\gamma_0\ \gamma_1 : | \GG |\}
    \to \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \to \mProp \\
    & \textsf{coe} : \{\gamma_0\ \gamma_1 : | \GG |\} \to \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \\
    & \dots
  \end{aligned}
}{
  A : \Ty\ \GG
}
\]

@@

An inductive-recursive universe is given by a type of codes $\U : \mType$, and a
family $\El : \U \to \mType$ that assigns, to each code corresponding to some
type, the meta-theoretic type of its elements. The resulting definition is
inductive-recursive because the inductive type of codes is defined
simultaneously with the recursive function $\El$.

@@

\begin{align*}
  & \textsf{data}\ \U\ : \mType \\
  & \quad \textsf{bool} : \U \\
  & \quad \textsf{pi} : (A : \U) \to (\El\ A \to \U) \to \U \\
  & \El\ \textsf{bool} = \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B) = (a : \El\ A) \to \El\ (B\ a)
\end{align*}

%% Note that $\U$ is defined to be in $\mType_1$ rather than $\mType$. This is
%% because... \todo{TODO add explanation}.

Induction-recursion is arguably the nicest and more natural way to define
internal universes in Type Theory, however it is not always required.
%
We can translate basic instances of induction-recursion into inductive families
using the equivalence of $I$-indexed families of types and types over $I$
(that is, $\mType_i \to I$).

@@

Note that this construction gives rise to a universe in $\mType_1$, rather than
$\mType$, since the definition of $\U$ quantifies over all possible sets. This
also means that a metatheory with at least one universe is required for this
construction.

@@

To construct the universe of setoids in the setoid model, we first of all need
to define a type $\Uty : \Ty\ \GG$ for every $\GG : \Con$, and for every $A :
\Tm\ \GG\ \Uty$ a type $\mathds{E}l\ A : \Ty\ \GG$. Recalling Section TODO,
these are essentially record types made of several components. Since $\Uty$ is a
closed type, it requires the same data of a setoid; in particular, we need a type
of codes together with an equivalence relation reflecting equality of codes:
%
\begin{align*}
  & \textsf{data} \ \setoidU : \mType_1 \\
  & \_\sim_\setoidU\_ : \setoidU \to \setoidU \to \mProp_1
  %% & \El : \setoidU \to \Set \\
  %% & \_\vdash\_\sim_\El\_ : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \Prop
\end{align*}
%
in addition to proofs that these are indeed equivalence relations
%
\begin{align*}
  & \textsf{refl}_\setoidU : (A : \setoidU) \to \equ{A}{A} \\
  & \textsf{sym}_\setoidU : \equ{A}{B} \to \equ{B}{A} \\
  & \textsf{trans}_\setoidU : \equ{A}{B} \to \equ{B}{C} \to \equ{A}{C}
\end{align*}

@@

Note that $\eqel{\_}{\_}{\_}$ is indexed over equality on the universe, because
$\El$ is a setoid morphism, hence in particular it must respect the setoid
equality of $\setoidU$.
%
We also require data and proofs that make sure we get setoids out of $\El$:

@@

[todo... explanation]
%
\begin{align*}
  & | \Uty | :\equiv \lambda\, \gamma . \, \setoidU \\
  & \Uty^\sim :\equiv \lambda \,p\, x\, y . \,x \sim_{\setoidU} y \\
  & \textsf{refl}\ \Uty :\equiv \reflu \\
  & ... \\
  & \textsf{coe}\ \Uty :\equiv \lambda\,p\,a . \, a \\
  & \textsf{coh}\ \Uty :\equiv \lambda\,p . \, \reflu
\end{align*}

Given $A : \Tm\ \GG\ \Uty$, we define $\EL\ A$ as follows:
%
\begin{align*}
  & | \EL\ A | :\equiv \lambda \, \gamma . \, \El\ (| A | \,\gamma) \\
  & (\EL\ A)^\sim :\equiv \lambda \, p \, x \, y . \, \eqel{A^\sim\,p}{x}{y} \\
  & \textsf{refl}\ (\EL\ A) :\equiv \reflel \\
  & ... \\
  & \textsf{coe}\ (\EL\ A) :\equiv \lambda\,p.\, \textsf{coe}_{\El}\,(A^\sim\,p) \\
  & \textsf{coh}\ (\EL\ A) :\equiv \lambda\,p.\, \textsf{coh}_{\El}\,(A^\sim\,p)
\end{align*}

We can show that $\Uty$ is closed under $\Pi$ types and \textsf{Bool}, including
$\beta$ and $\eta$ laws. Moreover, all the necessary equations hold
definitionally in the metatheory.

This gives a complete definition of a universe of setoids. However, as already
pointed out, this definition uses induction-recursion. Moreover, the kind of
recursion involved in this definition is particularly complex, since it involves
several recursive definitions whose types contain mutually defined functions. It
is not clear whether this definition schema, which we could call
induction-recursion-recursion, can be reduced to simpler, better understood
notions of induction-recursion like in [dybjer-setzer]. Nevertheless, even the
need to extend the metatheory with simple forms of induction-recursion goes
against our objective to keep the metatheory as small and essential as possible.

The goal of the next section is therefore to transform the
inductive-recursvie-recursive definition ourlined above to one that doesn't use
induction-recursion.
%
This transformation is inspired by the well-known trick to eliminate
induction-recursion described in Section X, but modified in a novel way to
account for the presence of $\mProp$-valued types. To our knowledge, this is the
first time this reduction trick is applied to an inductive-recursive type of
this kind.

@@

The reduction of the IR universe to induction works, conceptually, as in the
small example of Section X.
%
However, in addition to $\ad{in-U}$ for defining $\U$, we also introduce a
family $\ad{in-U∼}$ of binary relations between types in the universe, from
which we then define $\_\sim_{\setoidU}\_$.
%
\begin{align*}
  & \textsf{data}\ \ad{in-U} : \Set \to \Set_1 \\
  & \textsf{data}\ \ad{in-U∼} : \{A\ A' : \Set\} \to \ad{in-U}\ A \to \ad{in-U}\ A' \to (A \to A' \to \Prop) \to \Set_1
\end{align*}

Note that $\ad{in-U}$ appears in the sort of $\ad{in-U∼}$. Since these types are
mutually defined, this is an instance of \emph{induction-induction}, a schema
that allows the definition of a type mutually with other types that contain the
first one in their signature [cit]. The main example of induction-induction can
be found in intrinsic definitions of type theory in type theory [cite tt in tt].

@@

The full definition of the universe is concluded with the remaining definitions,
like $\reflu, \reflel$, etc., which can be adapted from their IR counterparts
more or less straighforwardly. The final result does not use
induction-recursion, but it's nevertheless an instance of infinitary
induction-induction. The ability to define arbitrary, infinitary
inductive-inductive types clashes, again, with our objective of keeping the
metatheory as small and simple as possible. The next step is therefore to reduce
this inductive-inductive universe to one that does not require (infinitary)
induction-induction.
